---
title: "Creating Plots- Layer by Layer"
subtitle: "Example Plots"
author: "Dr Abhijit Pakhare and Dr Ashwini Kalantri"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, hygge, ninjutsu]
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r, include = F}
knitr::opts_chunk$set(fig.width = 7, message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(ggrepel)
library(tidytext)
library(ggcharts)
library(colorspace)
library(ggforce)
```

```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}
@media print {
  .has-continuation {
    display: block;
  }
}
```



```{r data_import, echo=FALSE}
# Import key NFHS indicator dataset

df_nfhs<- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTdQq1FMM8HedkjOcxKEEqcTpei6RnYqIdbfgCilRk1XiPsloqaCvrJ-C1sLDKmYi3KJI6Tb9ga6xxZ/pub?gid=0&single=true&output=csv")

# Import GDP dataset

df_gdp<- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTdQq1FMM8HedkjOcxKEEqcTpei6RnYqIdbfgCilRk1XiPsloqaCvrJ-C1sLDKmYi3KJI6Tb9ga6xxZ/pub?gid=1296486578&single=true&output=csv")

df_nfhs<-df_nfhs %>% 
  mutate(Source=as_factor(Source))

df_nfhs_5<- df_nfhs %>% 
  filter(Year==2020) %>% 
  left_join(df_gdp, by=c("State","State.Code")) %>% 
  pivot_wider(names_from = Indicator,
              values_from = Estimate) %>% 
  mutate(IMR_Level= if_else(IMR>=30, ">=30", "<30"),
         State_Population_Cat= if_else(Population_Lakhs >=300, ">= 3 Cr", "<3 Cr")) %>% 
  mutate(IMR_Level=as_factor(IMR_Level)) 
```



# Visualizations to be demonstrated

--

Bubble diagram or Scatter plot showing correlation of Infant Mortality Rate (IMR) as per NFHS-5 data and Gross Domestic Product (State Net Domestic Product for the year 2019)

--

Box plot showing distribution of GDP per capita among states with low and high infant mortality (<30 per 1000 liver births and > 30 per 1000 live births)

--

Bar Plot showing selected indicators over years stratified by State

--

Line graph showing change in given indicators from NFHS-2 to NFHS-5 stratified by State

--

Dumbbell plot showing change in given indicators from NFHS-4 to NFHS-5 stratified by State


---
class: inverse, middle, center


# Scatter plot - Infant Mortality and GDP per capita





<!-- get a flipbook version of the bubble code chunk - pauses are where parentheses are balanced -->
---

`r chunk_reveal("scatter", title="Scatter Plot")`

```{r scatter, include = FALSE}
  ggplot(data = df_nfhs_5)+
  aes(x=GDP_Per_Capita_2019_in_thousands) +
  aes(y=IMR) +
  geom_point()
```
---


## Scatter plot- Infant Mortality and GDP per capita 

--
- Three critical layers 

--
- Data - indicated by name of dataframe 
  
--
- Mapping of aesthetic `aes()` 
  
--
- Geometry to visualize `geom_point()` 
  
--
- Further customization is needed 

--
- Let's do it!


---

`r chunk_reveal("bubble", title="Bubble Diagram")`

```{r bubble, include = FALSE}
df_nfhs_5 %>% 
  ggplot()+
  aes(x=GDP_Per_Capita_2019_in_thousands) +
  labs(x="GDP per capita")+
  aes(y=IMR) +
  geom_point(aes(colour=IMR))+
  scale_color_gradient2()+
  geom_text_repel(aes(label=State.Code))+
  aes(size=Population_Lakhs)+
  theme_minimal()+
  theme(legend.position = "bottom")+
  labs(title = "State wise Infant Mortality Rate and GDP per capita")+
  labs(caption = "Source: State Level Facts Sheets from NFHS-5, GDP Data- MOPSI Website")+
  theme(plot.title = element_text(face = "bold", 
                                  colour = "purple"))
```
---
### Using Log scale alters visualization 
```{r bubble_limit, include=TRUE, echo=FALSE}
p1<-df_nfhs_5 %>% 
  ggplot()+
  aes(x=GDP_Per_Capita_2019_in_thousands) +
  labs(x="GDP per capita")+
  aes(y=IMR) +
  geom_point(aes(colour=IMR), size=4)+
  scale_color_gradient2()+
  geom_text_repel(aes(label=State.Code))+
  theme_light()+
  theme(legend.position = "none")+
  labs(title = "State wise Infant Mortality Rate and GDP per capita")+
  labs(subtitle = "X-axis not altered")
```
```{r bubble_log, include=TRUE, echo=FALSE}
p2<-df_nfhs_5 %>% 
  ggplot()+
  aes(x=GDP_Per_Capita_2019_in_thousands) +
  labs(x="GDP per capita (Scale is Logarithmic)")+
  scale_x_log10()+
  aes(y=IMR) +
  geom_point(aes(colour=IMR), size=4)+
  scale_color_gradient2()+
  geom_text_repel(aes(label=State.Code))+
  theme_light()+
  theme(legend.position = "none")+
  labs(title = "State wise Infant Mortality Rate and GDP per capita")+
  labs(subtitle = "X-axis Log transformed")
```
```{r bubble_merge, echo=FALSE, include=TRUE, fig.width=14, fig.height=8}
library(patchwork)
p<-p1+p2
p
```


---
class: inverse, middle, center


# Box plot showing distribution of GDP per capita among states grouped by different levels of IMR





<!-- get a flipbook version of the box code chunk - pauses are where parentheses are balanced -->
---


`r chunk_reveal("box", title="Box Plot")`

```{r box, include = FALSE}
df_nfhs_5 %>% 
  ggplot()+
  aes(y=GDP_Per_Capita_2019_in_thousands) +
  labs(y="GDP per capita")+
  aes(x=IMR_Level) +
  geom_boxplot()+
  theme_minimal()+
  aes(fill=IMR_Level)+
  theme(legend.position = "bottom")+
  labs(title = "Box plot showing distribution of GDP per capita")
```

---
class: inverse, middle, center


# Bar plot showing indicators

---
`r chunk_reveal("bar1", title="Bar Chart")`




```{r bar1, include = FALSE}
df_nfhs %>% 
  filter(Indicator=="Stunting_U5") %>% 
  ggplot()+
  aes(x=Estimate)+
  aes(y=State.Code)+
  geom_col() +
  facet_wrap(facets = "Year")+
  facet_wrap(facets = "Year", scales = "free_y")+
  labs(x="Prevalence of Stunting")+
  theme_minimal()
```
---


# Were you satisfied with the previous bar chart? 

--
  - What was wrong? 

--
  - You need to scratch your head to understand anything 
  
--
  - Ordering states based on estimate will bring some clarity 

--
  - Let's re-do this chart 
---
`r chunk_reveal("bar", title="Bar Chart")`




```{r bar, include = FALSE}
df_nfhs %>% 
  filter(Indicator=="Stunting_U5") %>% 
  ggplot()+
  facet_wrap(~Year,scales = "free_y")+
  aes(y=reorder_within(State.Code,Estimate, Year))+
  scale_y_reordered()+
  aes(x=Estimate)+
  labs(x="Prevalence of Stunting")+
  geom_col(aes(fill=Estimate), show.legend = FALSE) +
  scale_fill_gradient2()+
  geom_text(aes(label=round(Estimate,1)), hjust=1, colour="white")+
  theme_minimal()+
  labs(title = "State wise prevalence of Stunting among children under 5 years of age")+
  labs(subtitle = "National Family Health Survey NFHS-3,4 and 5")
```

---
class: inverse, middle, center

# Lollipop chart 

---




```{r lollipop, include = TRUE, echo=FALSE, fig.width=14, fig.height=8}
  df_nfhs %>% 
  filter(Indicator=="Stunting_U5") %>% 
  ggplot()+
  aes(x=reorder_within(State.Code,Estimate, Year))+
  labs(x=NULL)+
  aes(y=Estimate)+
  labs(y="Prevalence of Stunting")+
  geom_segment(aes(xend=reorder_within(State.Code,Estimate, Year), yend=0)) +
  scale_x_reordered()+
  geom_point(aes(colour=Estimate), size=6, show.legend = FALSE)+
  scale_colour_gradient2()+
  geom_text(aes(label=round(Estimate,1)),  size=2, colour="white")+
  coord_flip()+
  facet_wrap(~Year,scales = "free_y")+
  theme_minimal()+
  labs(title = "State wise prevalence of stunting among children under 5 years of age")+
  labs(subtitle = "National Family Health Survey NFHS-3,4 and 5")

```
---
class: inverse, middle, center

# Line Diagram


---
`r chunk_reveal("line_diagram", title="Line Diagram")`





```{r line_diagram, include = FALSE}
df_nfhs %>% 
  filter(Indicator=="Stunting_U5" |Indicator=="Wasting_U5" | Indicator=="Underweight_U5") %>% 
  filter(State.Code !="AN" & State.Code !="LD" & State.Code !="CH" & State.Code !="DN"  & State.Code !="DD" & State.Code!="LH" & State.Code!="PY") %>% 
  ggplot()+
  aes(x=Year)+
  aes(y=Estimate)+
  labs(y=NULL)+
  geom_line()+
  aes(colour=Indicator)+
  facet_wrap(~State)+
  geom_text(aes(label=round(Estimate,0)),size=2)+
  theme_light()+
  theme(legend.position = "bottom")

```
---
class: inverse, middle, center

# Line Diagram (After editing)
---


```{r line_stunting, echo=FALSE, fig.width=15, fig.height=10}
line_graph<- df_nfhs %>% 
  filter(Indicator=="Stunting_U5" |Indicator=="Wasting_U5" | Indicator=="Underweight_U5") %>% 
  filter(State.Code !="AN" & State.Code !="LD" & State.Code !="CH" & State.Code !="DN"  & State.Code !="DD" & State.Code!="LH" & State.Code!="PY") %>% 
  ggplot()+
  aes(x=Year)+
  aes(y=Estimate)+
  labs(y=NULL)+
  geom_line(aes(colour=Indicator, alpha=1), size=1.5)+
  geom_point(aes(colour=Indicator, alpha=1), size=4)+
  geom_text_repel(aes(label=round(Estimate,0)),size=2)+
  facet_wrap(~State)+
  theme_light()+
  theme(legend.position = "bottom")+
  labs(title = "State wise prevalence of stunting, wasting and underweight among children below 5 years of age",
       subtitle = "Source- NFHS-3,4 and 5 State Facts Sheets")
line_graph
```
---
class: inverse, middle, center

# Difference in Wasting
---


```{r df_difference, echo=FALSE}
df_wasting_difference<- df_nfhs %>% 
  filter(Indicator=="Wasting_U5") %>% 
  select(Source,
         State,
         Estimate) %>% 
  pivot_wider(names_from = Source,
              values_from = Estimate) %>% 
  mutate(Difference=(`NFHS-5`-`NFHS-4`),
         Difference_Cat=if_else(Difference>0, "Worsened", "Improved")) %>% 
  drop_na()
df_wasting_difference
```

---
class: middle, center
## Diverging bar chart from `ggcharts` package
```{r divergingbar, echo=FALSE, fig.width=9, fig.height=6}
diverging_bar_chart(
  data=df_wasting_difference, 
  x=State, 
  y=Difference,
  bar_colors = c("#b32134", "#006400"),
  text_color = c("#b32134", "#006400"))+
  labs(title= "Difference in prevalence of wasting between NFHS-5 and NFHS-4",
    y="Difference",
    x=NULL,
    caption = "Chart created by using `divergeing_bar_chart` function of `ggcharts` package")
```
---
## Code for Diverging bar chart from `ggcharts` package
```{r divergingbar1, echo=TRUE, eval=FALSE}
library(ggcharts)
# One line code will generate default plot
diverging_bar_chart(data=df_wasting_difference, x=State, y=Difference)

# Further customization
diverging_bar_chart(
  data=df_wasting_difference, 
  x=State, 
  y=Difference,
  bar_colors = c("#b32134", "#006400"),
  text_color = c("#b32134", "#006400"))+
  labs(title= "Difference in prevalence of wasting between NFHS-5 and NFHS-4",
    y="Difference",
    x=NULL,
    caption = "Chart created by using `divergeing_bar_chart` function of `ggcharts` package")
```

---
class: middle, center

## Dumbbel plot `ggcharts` package
```{r dumbell, echo=FALSE, fig.width=9, fig.height=6}
dumbbell_chart(data=df_wasting_difference, x=State, y1=`NFHS-5`, y2=`NFHS-4`)+
  labs(title="Prevalence of wasting across states, NFHS-4 & 5",
       y="Prevalence of Wasting",
       x=NULL,
       caption = "Chart created by using `dumbbell_chart` function of `ggcharts` package")
```
---
## Code for Dumbbel plot `ggcharts` package
```{r dumbell1, echo=TRUE, eval=FALSE}
library(ggcharts)
# One line code will generate default plot
dumbbell_chart(data=df_wasting_difference, x=State, y1=`NFHS-5`, y2=`NFHS-4`)

# Further labels, title etc

dumbbell_chart(data=df_wasting_difference, x=State, y1=`NFHS-5`, y2=`NFHS-4`)+
  labs(title="Prevalence of wasting across states, NFHS-4 & 5",
       y="Prevalence of Wasting",
       x=NULL,
       caption = "Chart created by using `dumbbell_chart` function of `ggcharts` package")
```





