---
#############################################################
#                                                           #
# Click on "Run Document" in RStudio to run this worksheet. #
#                                                           #
#############################################################
title: "Public Health Data Visualization for Beginers Exercise-2"
author: "Abhijit P Pakhare and Ashwini Kalantri"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggrepel)
library(lubridate)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE, comment = "")

# Import key NFHS indicator dataset

df_nfhs<- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTdQq1FMM8HedkjOcxKEEqcTpei6RnYqIdbfgCilRk1XiPsloqaCvrJ-C1sLDKmYi3KJI6Tb9ga6xxZ/pub?gid=0&single=true&output=csv")

# Import GDP dataset

df_gdp<- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTdQq1FMM8HedkjOcxKEEqcTpei6RnYqIdbfgCilRk1XiPsloqaCvrJ-C1sLDKmYi3KJI6Tb9ga6xxZ/pub?gid=1296486578&single=true&output=csv")

df_nfhs<-df_nfhs %>% 
  mutate(Source=as_factor(Source))

df_nfhs_5<- df_nfhs %>% 
  filter(Year==2020) %>% 
  left_join(df_gdp, by=c("State","State.Code")) %>% 
  pivot_wider(names_from = Indicator,
              values_from = Estimate) %>% 
  mutate(IMR_Level= if_else(IMR>=30, ">=30", "<30")) %>% 
  mutate(IMR_Level=as_factor(IMR_Level)) %>% 
  filter(GDP_Per_Capita_2019_in_thousands>0) %>% 
  rename(GDP_PC_thousands=GDP_Per_Capita_2019_in_thousands) %>% 
  select(State.Code,
         Population_Lakhs,
         GDP_PC_thousands,
         IMR,
         Immunization_Full,
         Immunization_BCG,
         State)
```
**Attribution- This exercise is adapeted and modified from course material posted by Claus O. Wilke which is licensed under Creative Commons Attribution CC BY 4.0. https://github.com/wilkelab/SDS375 ** 

## Introduction

In this worksheet, we will learn how to create bar charts or column charts and its variants.

We will be using the R package **tidyverse**, which includes `ggplot()` and related functions.

```{r library-calls, echo = TRUE, eval = FALSE}
# load required library
library(tidyverse)
```

The dataset we will be working with contains the key indicators published in [State Fact Sheets](http://rchiips.org/nfhs/factsheet_NFHS-5.shtml) of National Family Health Surveys (2-5) :
```{r nfhs_dataset, echo = TRUE}
df_nfhs
```
For this exercise we will focus on indicator of stunting among children under 5 years of age. So we will filter only relevant rows containing information on stunting. This is done by using `filter` function of `dplyr` package which is part of `tidyverse` package family. Indicator column needs to be filtered for selecting rows containing "Stunting_U5".

```{r select_stunting, echo=TRUE}
df_nfhs_stunting<- df_nfhs %>% 
  filter(Indicator=="Stunting_U5")
df_nfhs_stunting
```

## Basic Column Chart

In the most basic use of bar chart, we call the `ggplot()` function with a dataset and an aesthetic mapping (created with `aes()`), and then we add a geom, such as `geom_col()` to draw columns.
### Data and axis mapping
Try this for yourself. Map the column `State.Code` onto the x axis and the column `Estimate` onto the y axis.  

```{r ggplot_blank, exercise=TRUE}
ggplot(df_nfhs_stunting, aes(x = ___, y = ___)) 
```

```{r ggplot_blank-hint}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = ___)) 
```

```{r ggplot_blank-solution}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) 
```

Axes are mapped, now  use `geom_bar()` to display the data.

```{r ggplot, exercise=TRUE}
ggplot(df_nfhs_stunting, aes(x = ___, y = ___)) +
  ___()
```

```{r ggplot-hint}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom____()
```

```{r ggplot-solution}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col()
```
If you carefully look at the graph, you will know something is wrong. Dataset contains information on various rounds of survey and this chart has simply added all values for the respective state. So let us map NFHS Survey round number indicated by variable "Source" to `fill` aesthetic and see what happens.
```{r ggplot_fill, exercise=TRUE}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col()
```

```{r ggplot_fill-hint}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=___))
```

```{r ggplot_fill-solution}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=Source))
```
And now swap which column you map to x and which to y. This can be done by changing aesthetic mapping in `aes` or by flipping coordinates using `coord_flip` function.

```{r ggplot3, exercise=TRUE}
ggplot(df_nfhs_stunting, aes(x = ___, y = ___)) +
  geom_col(aes(fill=Source))
```

```{r ggplot3-solution}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=Source))+
  coord_flip()
```
This looks better than earlier but it is difficult to read. NFHS round wise prevalence can not be directly visualized. What needs to be done here is creating panels by NFHS round. This is called as facetting. Let us learn it in next section.

## Facetting
We can add facets in `ggplot2` by adding a layer of `facet_wrap` function when we have only one facetting variable, like in our case we need to facet by "Source". There is `facet_grid` function in which we can fact by two variables. Let us add `facet_wrap` layer to the previous plot and assign facets="Source".

```{r ggplot_facets, exercise=TRUE}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=Source))+
  coord_flip()+
  facet_wrap(facets = ___)
```

```{r ggplot_facets-solution}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=Source))+
  coord_flip()+
  facet_wrap(facets = "Source")

```
Oh...There are many missing values owing to partial results of NFHS-5 and some states are created recently. We thus need to restrict our graph to non-missing values in factes. This is done by setting `scales` to `free_x` or `free_y`. Since we have flipped the coordinates before, we will set scales="free_y" under `facet_wrap` function.

```{r ggplot_facets1, exercise=TRUE}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=Source))+
  coord_flip()+
  facet_wrap(facets = "Source", scales = ___)

```

```{r ggplot_facets1-solution}
ggplot(df_nfhs_stunting, aes(x = State.Code, y = Estimate)) +
  geom_col(aes(fill=Source))+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")

```
You don't seem to be happy yet. We can understand you need to scratch your head to understand which states had higher prevalence of stunting in each survey. You may want to arrange bars in descending order. Now this is becoming bit complicated but there is always a solution in **R**. We will need to load `tidytext` package which gives function for reordering withing facets `reorder_within` and `scale_y_reordered` or `scale_x_reordered`. We have to specify three variables in `reorder_within` function, first is variable to reordered (in our case "State.Code"), then column values to be reordered (in our case "Estimate" column) and facetting variable (in our case "Source"). Let us map variable in `aes` by using `reorder_within` and then add `scale_x_reordered` layer.

```{r ggplot_facets1_reorder, exercise=TRUE}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(___, ___, ___), y = Estimate)) +
  geom_col(aes(fill=Source))+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")

```
```{r ggplot_facets1_reorder-solution}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(State.Code, Estimate, Source), y = Estimate)) +
  geom_col(aes(fill=Source))+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")

```



## Changing Theme

You may not like the default plot produced in earlier examples. `ggplot` provides options for customizations however it may need further travel on learning curve. We will now learn use of some themes available within `ggplot2` package. One of the popular theme is `theme_minimal` . Try adding it as a layer in the plot produced in previous example.

```{r ggplot-themes, exercise=TRUE}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(State.Code, Estimate, Source), y = Estimate)) +
  geom_col(aes(fill=Source))+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")+
  ___
```


```{r ggplot-themes-solution}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(State.Code, Estimate, Source), y = Estimate)) +
  geom_col(aes(fill=Source))+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")+
  theme_minimal()
```

## Adding titles, subtitles, labels and captions
Data visualization should be standalone and self-explanatory. Therefore we need to modify our plot and add titles, subtitles, axis labels and captions. In `ggplot2` we can add these as a layer. Let us modify our code and assign titles. We have use `labs()` layer for this purpose.  

Assign,  

title="Prevalence of stunting among Under-5 childrenin India",  

subtitle="Data Source- NFHS-3, 4 and 5 Fact Sheets",  

caption="Multi-panel diagram showing prevalence of stunting among children below 5 years of age. \n Within each panel states are ranked in descending order of prevalence",  


x="State/UT"  

y="Prevalence of Stunting",  


```{r ggplot-titles, exercise=TRUE}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(State.Code, Estimate, Source), y = Estimate)) +
  geom_col(aes(fill=Source))+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")+
  theme_minimal() +
  ___
```



```{r ggplot-titles-solution, fig.width=10, fig.height=8}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(State.Code, Estimate, Source), y = Estimate)) +
  geom_col(aes(fill=Source))+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")+
  theme_minimal()+
  labs(title="Prevalence of stunting among Under-5 childrenin India",
subtitle="Data Source- NFHS-3, 4 and 5 Fact Sheets",
caption="Multi-panel diagram showing prevalence of stunting among children below 5 years of age. \n Within each panel states are ranked in descending order of prevalence",
x="State/UT",
y="Prevalence of Stunting")+
  theme(legend.position="bottom")
```
## Shortcut 
R users are fantastic and they want to ease the burden of writing functions for other users. We can create the chart in very minimal code by using `ggcjarts` package. Code and output is displayed below. This chart can be further customized by using `ggplot2` functions/layers.

```{r ggchart_demo, echo=TRUE, fig.width=8, fig.height=8}
library(ggcharts)
df_nfhs_stunting %>% 
  bar_chart(x=State.Code, y=Estimate, facet=Source)
```

## Bar Chart Variant  

Length of bar sometimes distract our attention. We can just plot the point and label it with an estimate. For this we need to use `geom_point` instead of `geom_col`. This plot shows the same.
```{r ggplot-point-solution, echo=TRUE, fig.width=8, fig.height=8}
library(tidytext)
ggplot(df_nfhs_stunting, aes(x = reorder_within(State.Code, Estimate, Source), y = Estimate)) +
  geom_point(size=7)+
  geom_text(aes(label=round(Estimate,1)), size=2.5, colour="white")+
  scale_x_reordered()+
  coord_flip()+
  facet_wrap(facets = "Source", scales = "free_y")+
  theme_minimal()+
  labs(title="Prevalence of stunting among Under-5 childrenin India",
subtitle="Data Source- NFHS-3, 4 and 5 Fact Sheets",
caption="Multi-panel diagram showing prevalence of stunting among children below 5 years of age. \n Within each panel states are ranked in descending order of prevalence",
x="State/UT",
y="Prevalence of Stunting")+
  theme(legend.position="bottom")
```

